\emph{Gaia} DR2 adds clues for dynamical ejection of a $\sim$$150\,M_\odot$ star

### 2018-05-22 VFTS682.tex
minor in Sec. 2.2

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS 682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS 682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind obscuring the stellar surface. The error quoted above on the
radial velocity only include statistical uncertainties from the
spectral analysis, and not the systematic errors arising from the
modeling of emission line stars.

@@

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS 682. }
  \begin{tabular}{l|c|c}
  \tableline
    \tableline
    Parameter & Value & Source\\
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                      & \multirow{4}{*}{\emph{Gaia} DR2}\\[5pt]
    DEC \hfill [degree] & -69.07 % $\pm$  0.05
                      & \\[5pt]
    \cline{1-2}
    $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.12$\pm$0.07 & \\[5pt]
    $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.08 & \\[5pt]
    \hline
    $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
    $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                      0.2$\pm$0.1 &
    \\[5pt]
    \hline
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
    \tableline
  \end{tabular}
  \tablecomments{We do not list the error on the RA and DEC positions,
    which can be retrieved from the Gaia DR2 catalog and are of order
    $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$. We adopt the peculiar
    radial velocity calculated using the  NV $\lambda4944$ line, which is
    less sensitive to the wind velocity structure.}
    \end{center}
  \label{tab:vfts682}
\end{table}

The \emph{Gaia} Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}. VFTS 682 is identified with the source id 4657685637907503744 in the \emph{Gaia} DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. \emph{Gaia}
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the \emph{Gaia}
data for VFTS 682 are reliable. However, the effective temperature
reported in \emph{Gaia} DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the \emph{Gaia} DR2
catalog.

We retrieve for VFTS 682 the position in RA and DEC
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS 682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

We define a local standard frame of reference to derive the peculiar velocity
of VFTS 682 with respect to the cluster R136 by selecting from the \emph{Gaia} DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
We select all the stars in a target of 0.05 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require a G-band
magnitude brighter than 17, corresponding roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, and the errors on the proper
motion components to be smaller than 0.07\,$\mathrm{mas\
  yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
  yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
remove obvious foreground stars by requiring the absolute value of
the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
This selection yields a sample of 66 stars.% ,

We calculate the averaged proper motion components for the whole
region using 
\begin{equation}
  \label{eq:mean}
  \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
      \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
  \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
    \frac{1}{\Delta \mu_i^2}} \ \ ,
\end{equation}
where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
on the proper motion component reported by \emph{Gaia}. The sums run over
all the $N=66$ stars in our selected sample. We evaluate each proper motion
component separately. For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error bars on
the distance determination when converting proper motions into
physical velocities. The data retrieved, and the ipython notebook used for the analysis
presented here will be made available at \todo{git repo bitbucket?}.

\todo{nebular lines to compare velocity of the gas?}
\todo{fix numbers}

>>>>

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS 682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS 682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. \cite{bressert:12} also noted that VFTS
682 shows a significant radial velocity offset compared to the nebular
lines from the gas surrounding it, suggesting peculiar motion in the
line of sight direction. However, the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind obscuring the stellar surface. The errors quoted above on the
radial velocities include only statistical uncertainties from the
spectral analysis, and not the systematic errors arising from the
modeling of emission line stars.

@@

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS 682. }
  \begin{tabular}{l|c|c}
  \tableline
    \tableline
    Parameter & Value & Source\\
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 % $\pm$  0.03
                      & \multirow{4}{*}{\emph{Gaia} DR2}\\[5pt]
    DEC \hfill [degree] & -69.07 % $\pm$  0.05
                      & \\[5pt]
    \cline{1-2}
    $\delta\mu_\mathrm{RA}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.12$\pm$0.07 & \\[5pt]
    $\delta\mu_\mathrm{DEC}^{Gaia}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.08$\pm$0.08 & \\[5pt]
    \hline
    $\delta\mu_\mathrm{RA}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & 0.01$\pm$0.13 & \multirow{2}{*}{\cite{platais:18}}\\[5pt]
    $\delta\mu_\mathrm{DEC}^{HST}$  \hfill[$\mathrm{mas\ yr^{-1}}$] &
                                                                      0.2$\pm$0.1 &
    \\[5pt]
    \hline
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & 30$\pm$20 & \cite{bestenlehner:11}\\
    \tableline
  \end{tabular}
  \tablecomments{We do not list the error on the RA and DEC positions,
    which can be retrieved from the Gaia DR2 catalog and are of order
    $\sim$$10^{-2}\,\mathrm{mas\ yr^{-1}}$. We adopt the peculiar
    radial velocity calculated using the  NV $\lambda4944$ line, which is
    less sensitive to the wind velocity structure (see also \Secref{data:vfts683}).}
    \end{center}
  \label{tab:vfts682}
\end{table}

The second \emph{Gaia} Data Release (DR2), which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}. VFTS 682 is identified with the source id 4657685637907503744 in the \emph{Gaia} DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. \emph{Gaia}
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the \emph{Gaia}
data for VFTS 682 are reliable. However, the effective temperature
reported in \emph{Gaia} DR2, based on the spectral fitting around the
CaII triplet, is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative \citep[see, e.g.,][]{hogg:18}. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the \emph{Gaia} DR2
catalog.

We retrieve for VFTS 682 the position in RA and DEC
in the International Celestial Reference System \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS 682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

We follow \citet{lennon:18} to define a local frame of reference and derive the peculiar velocity
of VFTS 682 with respect to the cluster R136. Their selection (see
their Sec.~2.1) of 153
bright ($G<17$) stars around R136 with reliable astrometric data from
\emph{Gaia} DR2 yields $\mu_\mathrm{RA}=1.74\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and $\mu_\mathrm{DEC}=0.70\pm0.20\,\mathrm{mas\ yr^{-1}}$
for the components of the mean motion of the region projected on the sky.

% We define a local standard frame of reference to derive the peculiar velocity
% of VFTS 682 with respect to the cluster R136 by selecting from the \emph{Gaia} DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
% We select all the stars in a target of 0.05 degrees around R136
% (NGC2070) fulfilling the following criteria. First, we require a G-band
% magnitude brighter than 17, corresponding roughly to the
% completeness level of the VFTS survey \citep[here we implicitly assume
% G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
% \texttt{astrometric\_excess\_noise} $< 1$, and the errors on the proper
% motion components to be smaller than 0.07\,$\mathrm{mas\
%   yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
%   yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
% remove obvious foreground stars by requiring the absolute value of
% the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
% This selection yields a sample of 66 stars.% ,

% We calculate the averaged proper motion components for the whole
% region using 
% \begin{equation}
%   \label{eq:mean}
%   \langle \mu_i\rangle = \frac{\sum_\mathrm{stars}\frac{1}{\Delta
%       \mu_i^2}\mu_i}{\sum_\mathrm{stars} \frac{1}{\Delta \mu_i^2}} \ \ , \
%   \ \Delta \langle \mu_i\rangle = \frac{\sqrt{N}}{\sum_\mathrm{stars}
%     \frac{1}{\Delta \mu_i^2}} \ \ ,
% \end{equation}
% where $i = \mathrm{RA}, \mathrm{DEC}$, and $\Delta \mu_i$ is the error
% on the proper motion component reported by \emph{Gaia}. The sums run over
% all the $N=66$ stars in our selected sample. We evaluate each proper motion
% component separately.
For simplicity, throughout this study, we assume the same
distance of $50$\,kpc to the star \citep[][]{pietrzynski:13}, and to
the 30 Doradus region as a whole. We do not consider the error bars on
the distance determination when converting proper motions into
physical velocities. % The data retrieved, and the ipython notebook used for the analysis
% presented here will be made available at \todo{git repo bitbucket?}.
